### 2025-05-23 docs/article/Sections/introduction.tex
Chapter 2: add Spring MVC subsection

Modern web applications rely on different rendering strategies to optimize
performance, user experience, and \textit{scalability}. The two most dominant
approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
    Rendering (CSR)}. SSR generates HTML content on the server before sending it to
the client, resulting in a faster \textit{First Contentful Paint (FCP)}
\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
can increase server load and reduce \textit{throughput}\footnote{The number of
    requests the server can handle per second (RPS)} since each request requires
additional processing before responding. In contrast, CSR shifts the rendering
workload to the browser—the server initially sends a minimal HTML document with
JavaScript, which dynamically loads the page content. While CSR reduces the
server’s burden, it can lead to a slower \textit{FCP}, as users must wait for
JavaScript execution before meaningful content appears.

@@

Efficient thread utilization is crucial for low-thread servers. These servers
must handle many concurrent requests without being overwhelmed, which
necessitates asynchronous, non-blocking request handling
\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
database queries and API calls, servers can avoid blocking threads while
waiting for data, thereby maximizing throughput and responsiveness. To support
this non-blocking architecture, PSSR implementations require template engines
that are compatible with asynchronous data models and I/O. Some modern template
engines, such as HtmlFlow \cite{htmlflow} and Thymeleaf \cite{thymeleaf}, have
been designed with these capabilities in mind. However, many traditional
template engines—particularly those using external domain-specific languages
\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
\texttt{Iterable} for data processing. This blocking behavior forces server
threads to remain idle until the entire HTML output is ready, undermining the
performance benefits of non-blocking I/O and limiting scalability in
high-concurrency environments.

With the introduction of \textit{Virtual Threads} \cite{oracle-virtual-threads}
in Java 21, it is now possible to execute blocking I/O operations in a
scalable, lightweight manner. This capability allows traditional template
engines—often reliant on blocking interfaces—to operate efficiently in
high-concurrency, non-blocking environments without requiring complex
asynchronous programming models. As a result, Progressive Server-Side Rendering
(PSSR) can now be implemented using familiar HTML templates, 
simplifying development and improving maintainability.

>>>>

Modern web applications rely on different rendering strategies to optimize
performance, user experience, and \textit{scalability}. The two most dominant
approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
    Rendering (CSR)}. SSR generates HTML content on the server before sending it to
the client, resulting in a faster \textit{First Contentful Paint (FCP)}
~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
can increase server load and reduce \textit{throughput}\footnote{The number of
    requests the server can handle per second (RPS)} since each request requires
additional processing before responding. In contrast, CSR shifts the rendering
workload to the browser—the server initially sends a minimal HTML document with
JavaScript, which dynamically loads the page content. While CSR reduces the
server’s burden, it can lead to a slower \textit{FCP}, as users must wait for
JavaScript execution before meaningful content appears.

@@

Efficient thread utilization is crucial for low-thread servers. These servers
must handle many concurrent requests without being overwhelmed, which
necessitates asynchronous, non-blocking request handling
~\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
database queries and API calls, servers can avoid blocking threads while
waiting for data, thereby maximizing throughput and responsiveness. To support
this non-blocking architecture, PSSR implementations require template engines
that are compatible with asynchronous data models and I/O. Some modern template
engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
been designed with these capabilities in mind. However, many traditional
template engines—particularly those using external domain-specific languages
~\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
\texttt{Iterable} for data processing. This blocking behavior forces server
threads to remain idle until the entire HTML output is ready, undermining the
performance benefits of non-blocking I/O and limiting scalability in
high-concurrency environments.

With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
in Java 21, it is now possible to execute blocking I/O operations in a
scalable, lightweight manner. This capability allows traditional template
engines—often reliant on blocking interfaces—to operate efficiently in
high-concurrency, non-blocking environments without requiring complex
asynchronous programming models. As a result, Progressive Server-Side Rendering
(PSSR) can now be implemented using familiar HTML templates, 
simplifying development and improving maintainability.

### 2025-05-23 docs/article/Sections/state-of-the-art.tex
Chapter 2: add Spring MVC subsection

@@

		&\textbf{DSL idiom}
		&\textbf{Data Model APIs}
    &\shortstack{\textbf{Asynchronous}\\\textbf{Support}}
    &\shortstack{\textbf{Type}\\\textbf{Safety}}
    &\shortstack{\textbf{HTML}\\\textbf{Safety}}
		&\shortstack{\textbf{Progressive}\\\textbf{Rendering}}

@@

		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}

@@

		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{red}{\times}$}

@@

		&External DSL
		&\shortstack{\texttt{Iterable}\\\texttt{Publisher}}
		&\texttt{Publisher}$^{(1)}$
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}

@@

		&Nested Eager
		&All
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}

@@

    \shortstack{\textbf{Groovy }\\\textbf{MarkupBuilder}}
		&Nested Lazy
		&All
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}

@@

		&Chaining
		&All
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}

@@

		&Nested Lazy
		&All
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\textcolor{PineGreen}{\checkmark}$^{(2)}$
		&\large{$\textcolor{PineGreen}{\checkmark}$}

>>>>

Spring MVC is a classical, synchronous web framework built on the
thread-per-request model. When an HTTP request is received, it is handled by a
dedicated thread from the server’s thread pool (e.g., Tomcat, Jetty, or
Undertow). This thread is responsible for executing the full request lifecycle
synchronously, including invoking controllers, executing business logic, and
rendering the response view. Because the model is blocking, the thread remains
occupied throughout the request—even during I/O operations such as database
queries or external API calls—which can lead to performance bottlenecks under
high concurrency.

Controllers in Spring MVC typically return Java objects or models that are
processed by view resolvers in conjunction with template engines like JSP,
Thymeleaf, or FreeMarker. These templates generate the full HTML response in
one pass, which is then sent to the client only after all necessary data has
been gathered and rendered. By default, this architecture limits the ability to
perform \textit{Progressive Server-Side Rendering (PSSR)}, as traditional
rendering waits for all data before sending any output. This results in higher
latency for clients and less responsive page loads, particularly for
data-intensive views. However, Spring MVC provides a mechanism for response
streaming via the \texttt{StreamingResponseBody} interface, introduced in
Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
Spring writes directly to the \texttt{HttpServletResponse} output stream,
allowing developers to send parts of the response incrementally as they become
available. This is particularly useful for:
\begin{itemize}
  \item Sending large responses without buffering the entire output in memory.
  \item Writing dynamic HTML in fragments as data is fetched or computed.
  \item Reducing Time to First Byte (TTFB) and improving perceived performance.
\end{itemize}

A regular usage might look like the following:
\begin{lstlisting}[language=Java]
  @GetMapping("/stream")
  fun handleStream(): StreamingResponseBody? {
      return StreamingResponseBody { outputStream ->
          for (i in 0..9) {
              val htmlFragment = "<p>Chunk " + i + "</p>\n"
              outputStream.write(htmlFragment.toByteArray())
              outputStream.flush() // ensure partial response is sent
              Thread.sleep(500) // simulate delay
          }
      }
  }
\end{lstlisting}

In this example, HTML content is written and flushed in discrete chunks,
allowing the client to progressively render the response as it arrives, while
the server continues processing additional data. Although
\texttt{StreamingResponseBody} does not eliminate the blocking nature inherent
in Spring MVC—since the handling thread remains active during the streaming
process—it does enable a basic form of PSSR by facilitating incremental content
delivery. This approach is particularly useful for long-running operations or
large response bodies, where providing early partial output can enhance
perceived performance.

However, a key limitation we encountered with this method involves the servlet
response buffer. By default, most servlet containers (e.g., Tomcat) use a
response buffer size of approximately 8KB. Data written to the response output
stream is held in this buffer until it reaches capacity. Consequently, if the
total size of the initial HTML content is less than the buffer threshold, the
server will not transmit any data to the client until the buffer is filled.
This buffering behavior introduces a delay in sending the first chunk of
content, thereby negating the benefits of progressive rendering. In scenarios
where the rendered HTML template does not produce sufficient data to exceed the
buffer threshold early, PSSR becomes ineffective, and the client experiences a
delayed initial response.

@@

     & \textbf{DSL idiom}
     & \textbf{Data Model APIs}
     & \shortstack{\textbf{Asynchronous}           \\\textbf{Support}}
     & \shortstack{\textbf{Type}                   \\\textbf{Safety}}
     & \shortstack{\textbf{HTML}                   \\\textbf{Safety}}
     & \shortstack{\textbf{Progressive}            \\\textbf{Rendering}}

@@

     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}

@@

     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}

@@

     & External DSL
     & \shortstack{\texttt{Iterable}               \\\texttt{Publisher}}
     & \texttt{Publisher}$^{(1)}$
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}

@@

     & Nested Eager
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}

@@

    \shortstack{\textbf{Groovy }                   \\\textbf{MarkupBuilder}}
     & Nested Lazy
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}

@@

     & Chaining
     & All
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}

@@

     & Nested Lazy
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \textcolor{PineGreen}{\checkmark}$^{(2)}$
     & \large{$\textcolor{PineGreen}{\checkmark}$}
